---
title: "Using mixchisq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using mixchisq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`mixchisq` provides a small set of utilities for finite mixtures of chi-square distributions. The motivating use case is practical rather than theoretical: sometimes your reference distribution is not a single chi-square law, and you need a convenient way to evaluate its CDF, density, quantiles, or simulate from it.

The package supports both standard finite mixtures such as

\[
0.5\chi^2_2 + 0.5\chi^2_3,
\]

and chi-bar-square style laws with an atom at zero, such as

\[
0.5\chi^2_0 + 0.5\chi^2_1.
\]

```{r}
library(mixchisq)
```

## A simple mixture without an atom at zero

Consider the mixture

\[
X \sim 0.5\chi^2_2 + 0.5\chi^2_3.
\]

You can evaluate probabilities, densities, quantiles, and simulate from it directly.

```{r}
pmixchisq(5, df = c(2, 3), w = c(0.5, 0.5))
dmixchisq(5, df = c(2, 3), w = c(0.5, 0.5))
qmixchisq(0.95, df = c(2, 3), w = c(0.5, 0.5))
set.seed(123)
rmixchisq(10, df = c(2, 3), w = c(0.5, 0.5))
```

You can also plot the CDF or density.

```{r, fig.width=6, fig.height=4}
plot_mixchisq(df = c(2, 3), w = c(0.5, 0.5), type = "cdf")
```

## A chi-bar-square style mixture

Now consider

\[
X \sim 0.5\chi^2_0 + 0.5\chi^2_1,
\]

which has a point mass of size 0.5 at zero.

```{r}
pmixchisq(c(-1, 0, 0.2, 1), df = c(0, 1), w = c(0.5, 0.5))
qmixchisq(c(0.25, 0.50, 0.90), df = c(0, 1), w = c(0.5, 0.5))
set.seed(1)
rmixchisq(20, df = c(0, 1), w = c(0.5, 0.5))
```

The jump at zero is reflected in the CDF plot.

```{r, fig.width=6, fig.height=4}
plot_mixchisq(df = c(0, 1), w = c(0.5, 0.5), type = "cdf")
```

## Replacing ad hoc simulation code

A common pattern is to simulate a mixture using a latent Bernoulli draw and then generate from one of two chi-square distributions. For example:

```{r}
set.seed(123)
manual <- {
  z <- runif(10000)
  as.numeric(z < 0.5) * rchisq(10000, df = 2) +
    as.numeric(z > 0.5) * rchisq(10000, df = 3)
}
summary(manual)
```

With `mixchisqutils`, that becomes more direct and easier to read:

```{r}
set.seed(123)
auto <- rmixchisq(10000, df = c(2, 3), w = c(0.5, 0.5))
summary(auto)
```

And if you want the CDF on a grid, you do not need a large Monte Carlo approximation first.

```{r}
grid <- seq(0.1, 10, by = 0.1)
cdf_vals <- pmixchisq(grid, df = c(2, 3), w = c(0.5, 0.5))
head(cdf_vals)
```

## Typical workflow

A practical workflow looks like this:

1. identify the mixture weights and degrees of freedom suggested by your constrained test,
2. evaluate `pmixchisq()` for p-values or CDF comparisons,
3. use `qmixchisq()` for critical values,
4. use `rmixchisq()` for simulation studies or calibration checks,
5. use `plot_mixchisq()` to sanity-check the reference law visually.

## Summary

`mixchisq` is meant to be a compact utility package. It does not try to derive the mixture for you. Instead, once you know the mixture you want, it gives you a clean way to work with it in code.
